---
title: "Professional Football Statistics"
author: "Tenzin Gyaltsen, Shen Rothermel"
date: "2025-05-15"
output: html_document
runtime: shiny
---

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Libraries}
library(dplyr)
library(ggthemes)
library(janitor)
library(polite)
library(purrr)
library(readr)
library(rvest)
library(stringr)
library(tidyverse)
```

## Introduction

In this project, we explore soccer statistics from multiple professional football leagues using data from FBref (https://fbref.com/en/comps), a trusted site for advanced football analytics. While we initially focused on Major League Soccer (MLS), we extended our analysis to include other major international competitions such as the Premier League, La Liga, Bundesliga, and Serie A.

Our goal was to collect and organize standardized squad-level statistics across leagues to support comparative analysis. Specifically, we targeted the "Squad Standard Stats" tables on each competition’s main stats page. These tables contain information on team performance metrics such as matches played, goals, assists, average age, possession %, and more.

## Motivation

We chose this dataset primarily out of personal interest: one of us enjoys following global football news, while the other is an avid FC25 player. Beyond our curiosity, we recognized that this data offers a rich opportunity for cross-league comparisons.

By scraping the same type of statistics from each league, we aimed to answer questions such as:

-   Do older squads tend to score more or less?
-   Is there a relationship between average age and possession percentage?
-   How does team performance (e.g., goals, assists) vary across leagues?

These questions open the door for future data visualizations (like scatterplots or heatmaps) and statistical modeling (e.g., regression of goals on age or possession).

```{r Initial Information}
# For mapping use in scraping multiple data from multiple leagues
squad_leagues <- tibble::tibble(
	league = c("Bundesliga", "La Liga", "MLS", "Premier League", "Serie A"),
	# Each league link is almost identical, with a unique number at the end
	url_number = c(20, 12, 22, 9, 11),
	table_num = c(3, 3, 5, 3, 3) # Identifies where the table of interest is
)

# Same as above but for standings instead of squad statistics
standings_leagues <- tibble::tibble(
	league = c("Bundesliga", "La Liga", "MLS East",
						 "MLS West", "Premier League", "Serie A"),
	url_number = c(20, 12, 22, 22, 9, 11),
	table_num = c(1, 1, 1, 3, 1, 1)
)
```

```{r Custom Scrape Function}
scrape_fbref_table <-
	function(url_number, table_number, n_cols = 20, p_cols = 2) {
		# Every link starts the same, with a unique
		# number at the end corresponding to the league
		session <- bow(paste("https://fbref.com/en/comps/", url_number, sep = ""))
		
		# We are interested in just the tables, which contain data
		all_tables <- scrape(session) |> 
			html_nodes(css = "table") |> 
			html_table(fill = TRUE)
		
		# Specifically interested in just one table at a time
		clean_table <- all_tables[[table_number]]
		
		# Annoyingly, the very first table of MLS will be scraped in completely
		# tidy, but nothing else, so we have to check whether the column names
		# are tidy yet, and if not, then tidy it
		if (names(all_tables[[table_number]])[1] == "") {
			clean_table <- clean_table |> 
				row_to_names(1) |> 
				clean_names() |> 
				select(all_of(1:n_cols)) |> 
				mutate(across(all_of(p_cols:n_cols), parse_number))
		}
	
		return(clean_table)
	}
```

```{r Scrape Squad Standard Stats}
# Map each league's url number and table of interest to the scrape function
# then bind all resulting tables' rows together into a single tibble
squads_binded <- bind_rows(
	pmap(
		list(squad_leagues$url_number, squad_leagues$table_num),
		scrape_fbref_table
	),
	.id = "league"
) |> mutate(league = squad_leagues$league[as.integer(league)])
# Finish with replacing the identifier column with actual league names
```

```{r Squad Standard Stats Augmentation}
# Calculate some extra variables
squads <- squads_binded |> 
	mutate(goals_per_game = gls / mp,
				 assist_rate = ast / gls,
				 discipline_score = crd_y + 2 * crd_r,
				 min_per_player = min / number_pl)
```

```{r Scrape Standings Stats}
# Same deal as squads but for standings instead
standings_binded <- bind_rows(
	pmap(
		list(standings_leagues$url_number, standings_leagues$table_num, replicate(6, 15), replicate(6, 3)),
		scrape_fbref_table
	),
	.id = "league"
) |> mutate(league = standings_leagues$league[as.integer(league)])
```

```{r Standings Stats Augmentation}
standings <- standings_binded |>
	mutate(`Top Team Scorer Goals` =
				 		parse_number(str_extract(`Top Team Scorer`, "^.* - (\\d{1,2})$", 1)),
				 `Top Team Scorer` =
				 		str_extract(`Top Team Scorer`, "^(.*) - \\d{1,2}$", 1)) |>
	relocate(`Top Team Scorer Goals`, .after = `Top Team Scorer`)
```

```{r Avg Age vs Goals}
titlePanel("Team Average Age vs. Goals by League")

renderPlot({
	squads |> 
		ggplot() +
			geom_point(aes(x = age, y = gls, color = league, shape = league), size = 3) +
			labs(x = "Average Age",
					 y = "Goals",
					 color = "League",
					 shape = "League") +
			scale_color_colorblind()
})
```

```{r Possession vs Goals}
titlePanel("Possession vs. Goals by League")

renderPlot({
	squads |> 
		ggplot() +
			geom_point(aes(x = poss, y = gls, color = league), size = 2, show.legend = F) +
			facet_wrap(~league) +
			labs(x = "Possession",
					 y = "Goals") +
			scale_color_colorblind()
})
```

```{r Top Teams Goals per Game}
titlePanel("Top 5 Teams by Goals per Game by League")

renderPlot({
	squads |>
		group_by(league) |> 
		slice_max(goals_per_game, n = 5) |> 
		ggplot() +
			geom_col(aes(x = fct_reorder(squad, goals_per_game),
									 y = goals_per_game,
									 fill = league)) +
			labs(x = "Squad",
					 y = "Goals per Game",
					 fill = "League") +
			coord_flip() +
			scale_fill_colorblind()
})
```

```{r}
titlePanel("Standings")

inputPanel(
	checkboxInput("show_all", "Show All Rows", value = FALSE)
)

DT::renderDataTable({
	optionsStandings <- if (input$show_all) {
		list(bPaginate = FALSE)
	} else {
		list(
			pageLength = 5,
			lengthMenu = c(5, 10, 20, 40),
			rownames = FALSE
		)
	}
	
	DT::datatable(standings, options = optionsStandings)
})
```